---
title: "Introduction to NIHSnmr"
author: "Sergio Oller, Institute for Bioengineering of Catalonia"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to NIHSnmr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 5,
  collapse = TRUE,
  comment = "#>"
)
```

The `NIHSnmr` package was written with two purposes in mind:

- to help **data analysts and NMR scientists** to work with NMR samples.
- to help **IT pipeline builders** implement automated methods for preprocessing.

Functions from this package written for data analysts and NMR scientists are prefixed with
`nmr_`, while higher level functions written for IT pipeline builders are prefixed
with `pipe_`.

This vignette is written for the first group. It assumes some prior basic knowledge of
NMR and data analysis, as well as some basic R programming. In case you are interested in
building pipelines with this package, better check [Building Pipelines with NIHSnmr](building-pipelines-with-nihsnmr.html).

```{r}
library(NIHSnmr)
```


# Data: The `MeOH_serum_extraction` dataset

To explore the basics of the NIHSnmr package, we have included four NMR samples
in Bruker format bundled with the package. The samples are (c) Nestlé Institute of Health Sciences
and should not be redistributed without NIHS written consent, in case this package was ever published.

If you have installed this package, you can obtain the directory where the four samples
are with the command

```{r}
MeOH_serum_extraction_dir <- system.file("dataset-demo", package = "NIHSnmr")
print(MeOH_serum_extraction_dir)
```

The demo directory includes four samples (zipped) and a dummy Excel metadata file.

```{r}
fs::dir_ls(MeOH_serum_extraction_dir)
```

Given the name of the dataset, one may guess that the dataset was used to
check the Methanol extraction in serum samples. The dummy metadata consists of
dummy information, just for the sake of showing how this package can integrate
external metadata. The excel file consists of two tidy tables, in two sheets.

```{r}
MeOH_serum_extraction_xlsx <- file.path(MeOH_serum_extraction_dir, "dummy_metadata.xlsx")
print(readxl::read_excel(MeOH_serum_extraction_xlsx, sheet = 1))
print(readxl::read_excel(MeOH_serum_extraction_xlsx, sheet = 2))

```


# Loading samples

The function to read samples is called `nmr_read_samples`. It expects a character
vector with the samples to load that can be paths to directories of Bruker format
samples or paths to JDX files.

Additionally, this function can filter by pulse sequences (e.g. load only NOESY samples) or loading only metadata.

```{r load-samples}
zip_files <- fs::dir_ls(MeOH_serum_extraction_dir, glob = "*.zip")
print(zip_files)
dataset <- nmr_read_samples(sample_names = zip_files)
print(dataset)
```

# Adding metadata

Adding metadata is easy once we have the information in a dataset. The `nmr_meta_add`
allows us to append a data frame, using a common column for merging.

In this example we add the `SubjectID` and the `Age` columns. First we add the
`SubjectID` using the `NMRExperiment` as key and then we add the `Age` column
using the just added `SubjectID` column as key for merging.

```{r adding-metadata}
exp_subj_id <- readxl::read_excel(MeOH_serum_extraction_xlsx, sheet = 1)
subj_id_age <- readxl::read_excel(MeOH_serum_extraction_xlsx, sheet = 2)

dataset <- dataset %>%
  nmr_meta_add(metadata = exp_subj_id, by = "NMRExperiment") %>%
  nmr_meta_add(metadata = subj_id_age, by = "SubjectID")
```

We can retreive the table with the added metadata with:

```{r}
nmr_meta_get(dataset, groups = "external")
```

The dataset has several fields. You can check their definition with `?nmr_dataset`.

```{r}
print(names(dataset))
```

# Interpolation

1D NMR samples can be interpolated together, in order arrange all the spectra in
a matrix, with one row per sample. The main parameters we would need is the range
of ppm values that we want to interpolate and the resolution.

We can see the ppm resolution by looking at the ppm axis of one sample:

```{r}
ppm_res <- nmr_ppm_resolution(dataset)[[1]]
message("The ppm resolution is: ", format(ppm_res, digits = 2), " ppm")
```

We can interpolate the dataset, obtaining an `nmr_dataset_1D` object:

```{r}
dataset <- nmr_interpolate_1D(dataset, axis = c(min = 0.2, max = 10, by = 2.3E-4))
```

This operation changes the class of the object, as now the data is on a matrix. The
dataset is now of class `nmr_dataset_1D`. The `axis` element is now a numeric vector
and the `data_1r` element is a matrix.

# Plotting samples

The `NIHSnmr` package offers the possibility to plot `nmr_dataset_1D` objects.
Plotting many spectra with so many points is quite expensive so it is possible
to include only some regions of the spectra or plot only some samples.

Use `?plot.nmr_dataset_1D` to check the parameters, among them: 

- `NMRExperiment`: A character vector with the NMR experiments to plot
- `chemshift_range`: A ppm range to plot only a small region, or to reduce the resolution
- `interactive`: To make the plot interactive
- `...`: Can be used to pass additional parameters such as `color = "SubjectID"` that are
  passed as aesthetics to ggplot.

```{r}
plot(dataset, NMRExperiment = c("10", "30"), chemshift_range = c(2.2, 2.9))
```


# Exclude regions

Some regions can easily be excluded from the spectra with `nmr_exclude_region`.
Note that the regions are fully removed and not zeroed, as using zeros complicates
a lot the implementation^[e.g. it can inadvertedly distort the PQN normalization results]
and has little advantages.

```{r}
regions_to_exclude <- list(water = c(4.6, 5), methanol = c(3.33, 3.39))
dataset <- nmr_exclude_region(dataset, exclude = regions_to_exclude)
plot(dataset, chemshift_range = c(4.2, 5.5))
```

# Filter samples

Sometimes we just want to analyze a subset of the data, e.g. just a cohort. We
can filter some samples according to their metadata as follows:

```{r}
samples_10_20 <- filter(dataset, SubjectID == "Ana")
nmr_meta_get(samples_10_20, groups = "external")
```

# Robust PCA for outlier detection

The NIHSnmr package includes some robust PCA analysis for outlier detection.
With such a small demo dataset, it is not practical to use, but check out the
documentation of `nmr_pca_outliers_*` functions.

```{r}
pca_outliers_rob <- nmr_pca_outliers_robust(dataset, ncomp = 3)
nmr_pca_outliers_plot(dataset, pca_outliers_rob)
```

# Peak detection


```{r}

```

# Spectra alignment

# Normalization

# Peak integration


# Free experimentation

You can extract the raw matrix and manipulate it at will. As long as you don't
permute the rows, you can always replace the raw matrix of the `nmr_dataset_1D`
object:

```{r}
full_spectra_matrix <- nmr_data(dataset)
print(full_spectra_matrix[1:4, 1:6])
nmr_data(dataset) <- full_spectra_matrix # Rewrite the matrix
```


