---
title: "Plotting NMR samples"
author: "AlpsNMR authors"
package: AlpsNMR
abstract: >
  Plotting samples and data
date: "`r format(Sys.Date(), '%F')`"
output: "BiocStyle::pdf_document"
vignette: >
  %\VignetteIndexEntry{Vignette03: Plotting NMR samples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r}
library(ggplot2)
library(AlpsNMR)
```


(Vignette in progress)

# Creating interactive plots

Use `?plot.nmr_dataset_1D` to check the parameters, among them: 

- `NMRExperiment`: A character vector with the NMR experiments to plot 
- `chemshift_range`: A ppm range to plot only a small region, or to reduce the
resolution 
- `interactive`: To make the plot interactive - `...`: Can be used to
pass additional parameters such as `color = "SubjectID"` that are passed as
aesthetics to ggplot.

The option `interactive = TRUE` described above has some performance limitations.
As high performance workaround, you can make many plots interactive with the
function `plot_interactive`.

This function will use WebGL technologies to create a webpage that, once opened,
allows you to interact with the plot.

Due to technical limitations, these plots need to be opened manually and can't
be embedded in RMarkdown documents. Therefore, the function saves the plot in 
the directory for further exploration. Additionally, some old web browsers may
not be able to display these interactive plots correctly.

```
plt <- plot(dataset, NMRExperiment = c("10", "30"), chemshift_range = c(2.2, 2.8))
plot_interactive(plt, "plot_region.html")
```


# Session Information

```{r}
sessionInfo()
```
