---
title: "Introduction to AlpsNMR"
author: "AlpsNMR authors"
package: AlpsNMR
abstract: >
  An introduction to the AlpsNMR package, showing the most relevant functions and
  a proposed workflow. This includes loading bruker NMR samples, adding sample
  annotations, preprocessing the spectra, detecting outliers, detecting peaks,
  aligning the samples and integrating the peaks to build a peak table.
date: "`r format(Sys.Date(), '%F')`"
output:
  BiocStyle::pdf_document
vignette: >
  %\VignetteIndexEntry{Vignette 01: Introduction to AlpsNMR (start here)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Getting started

The `AlpsNMR` package has most of its functions prefixed with `nmr_`.
The main reason for this is to avoid conflicts with other packages.
Besides, it helps for autocompletion:
Most coding environments such as RStudio will let you see most of the function
names by typing `nmr_` followed by pressing the tab key.

This vignette assumes some basic knowledge of NMR and data analysis, and some
basic R programming.

We will start by loading `AlpsNMR` along some convenience packages:

```{r}
library(dplyr)
library(ggplot2)
library(readxl)
library(BiocParallel)
library(AlpsNMR)
```


# Enable parallellization

This package is able to parallellize several functions through the use of the
`BiocParallel` package. Whether to parallelize or not is left to the user
that can control the parallellization registering backends. Please check
`vignette("Introduction_To_BiocParallel", package = "BiocParallel")`.

```{r}
#register(SerialParam(), default = TRUE)  # disable parallellization
register(SnowParam(workers = 3), default = TRUE)  # enable parallellization with 3 workers
```


# Data: The `MeOH_plasma_extraction` dataset

To explore the basics of the AlpsNMR package, we have included four NMR samples
acquired in a 600 MHz Bruker instrument bundled with the package.
The samples are pooled quality control plasma samples, that were extracted with methanol.
They only contain small molecules.

If you have installed this package, you can obtain the directory where the
samples are with the command:

```{r}
MeOH_plasma_extraction_dir <- system.file("dataset-demo", package = "AlpsNMR")
MeOH_plasma_extraction_dir
```

The demo directory includes three zipped Bruker samples and a dummy Excel metadata file:

```{r}
list.files(MeOH_plasma_extraction_dir)
```

Since these are quality control samples, the metadata is a dummy table:

```{r}
MeOH_plasma_extraction_xlsx <- file.path(MeOH_plasma_extraction_dir, "dummy_metadata.xlsx")
annotations <- readxl::read_excel(MeOH_plasma_extraction_xlsx)
annotations
```

# Loading samples

The function to read samples is called `nmr_read_samples()`. It expects a
character vector with the samples to load that can be paths to directories of
Bruker format samples or paths to JDX files.

Additionally, this function can filter by pulse sequences (e.g. load only NOESY
samples) or loading only metadata.

```{r load-samples}
zip_files <- fs::dir_ls(MeOH_plasma_extraction_dir, glob = "*.zip")
zip_files
dataset <- nmr_read_samples(sample_names = zip_files)
dataset
```

If your samples happen to be in different **folders per class**, AlpsNMR provides
convenience functions to read them as well. With this example:

```
- your_dataset/
  + control/
     * 10/
     * 20/
     * 30/
  + mutated/
     * 10/
     * 20/
     * 30/
```

You could use:

```
dataset <- nmr_read_samples_dir(c("your_dataset/control", "your_dataset/mutated"))
dataset
```

If after reading the `?nmr_read_samples` page you still have issues, feel free
to open an issue at [https://github.com/sipss/AlpsNMR/issues] and ask for
clarification.

# Adding annotations

We can embed the external annotations we loaded above into the dataset:

```{r}
dataset <- nmr_meta_add(dataset, metadata = annotations, by = "NMRExperiment")
```

And retrieve them from the dataset:

```{r}
nmr_meta_get(dataset, groups = "external")
```


If you want to learn more about sample metadata (including acquisition and FID 
processing parameters), as well as more complex ways of adding annotations,
check out the `vignette("Vig02-handling-metadata-and-annotations", package = "AlpsNMR")`.


# Interpolation

1D NMR samples can be interpolated together, in order to arrange all the spectra 
into a matrix, with one row per sample. Here we choose the 
range of ppm values that we want to include in further analyses.

```{r}
dataset <- nmr_interpolate_1D(dataset, axis = c(min = -0.5, max = 10))
```

If the `axis = NULL` then the ppm axis is autodetected from the samples.

See `nmr_interpolate_1D()` for further reference on the axis options.

# Plotting samples

Plotting many spectra with so many points is quite expensive so it is possible
to include only some regions of the spectra or plot only some samples.

```{r}
plot(dataset, NMRExperiment = c("10", "30"), chemshift_range = c(2.2, 2.8))
```

See `vignette("Vig03-plotting-samples", package="AlpsNMR")` for more advanced plot options, or
having interactive plots.

# Exclude regions

Some regions can easily be excluded from the spectra with `nmr_exclude_region()`:

```{r}
regions_to_exclude <- list(water = c(4.6, 5), methanol = c(3.33, 3.39))
dataset <- nmr_exclude_region(dataset, exclude = regions_to_exclude)
plot(dataset, chemshift_range = c(4.2, 5.5))
```


# Filter samples

Maybe we just want to analyze a subset of the data, e.g., only a class group or
a particular gender. We can filter some samples according to their metadata as 
follows:

```{r}
samples_10_20 <- filter(dataset, SubjectID == "Ana")
nmr_meta_get(samples_10_20, groups = "external")
```


# Robust PCA for outlier detection

The AlpsNMR package includes robust PCA analysis for outlier detection.

```{r}
pca_outliers_rob <- nmr_pca_outliers_robust(dataset, ncomp = 3)
nmr_pca_outliers_plot(dataset, pca_outliers_rob)
```


# Baseline removal

Spectra may display an unstable baseline, specially when processing
blood/fecal samples. 

The peak detection and integration algorithms benefit from having
an estimation of the baseline, so it is advisable to compute it first
and check it fits as expected.


See before:

```{r}
plot(dataset, chemshift_range = c(3.5,3.8))
```

Estimate the baseline:

```{r}
dataset <- nmr_baseline_estimation(dataset, lambda = 9, p = 0.01)
```



And after:

```{r fig.height=10, fig.width=10}
# TODO: Simplify this plot
spectra_to_plot <- tidy(dataset, chemshift_range = c(3.5, 3.8), columns = "NMRExperiment")
baseline_to_plot <- tidy(dataset, chemshift_range = c(3.5, 3.8), columns = "NMRExperiment", matrix_name = "data_1r_baseline")

ggplot(mapping = aes(x = chemshift, y = intensity, color = NMRExperiment)) +
    geom_line(data = spectra_to_plot) +
    geom_line(data = baseline_to_plot, linetype = "dashed") + 
    facet_wrap(~NMRExperiment, ncol = 1)

```


# Peak detection

The peak detection is performed on short spectra segments using a continuous
wavelet transform. See `?nmr_detect_peaks` for more information.

Our current approach relies on the use of the baseline threshold 
(`baselineThresh`) automatically calculated (see `?nmr_baseline_threshold`) 
and the Signal to Noise Threshold (`SNR.Th`) to discriminate valid peaks 
from noise. 

The combination of the `baselineThresh` and the `SNR.Th` optimizes 
the number of actual peaks from noise.

The advantage of the `SNR.Th` method is that it estimates the noise
level on each spectra region independently, so in practice it can be used as
a dynamic baseline threshold level.

```{r}
baselineThresh <- nmr_baseline_threshold(dataset, range_without_peaks = c(9.5, 10))
nmr_baseline_threshold_plot(dataset, baselineThresh)
```


```{r}
peak_list <- nmr_detect_peaks(
    dataset,
    nDivRange_ppm = 0.1,
    scales = seq(1, 16, 2),
    baselineThresh = baselineThresh,
    SNR.Th = 3
)
```

We can get an overview of the number of peaks we detect on each sample and each chemical shift region:

```{r}
nmr_detect_peaks_plot_overview(peak_list)
```


We can explore in a more detailed way the detected peaks:

```{r}
nmr_detect_peaks_plot(dataset, peak_list, NMRExperiment = "10", chemshift_range = c(3, 3.3))
```

Let's the detected peaks in a smaller region across samples:

```{r}
peak_list_in_range <- filter(peak_list, ppm > 3.22, ppm < 3.24)
peak_list_in_range
```

```{r}
plot(dataset, chemshift_range = c(3.22, 3.25))
```


```{r dev="cairo_pdf"}
nmr_detect_peaks_plot_peaks(dataset, peak_list, peak_ids = peak_list_in_range$peak_id)

```


# Spectra alignment

To align the sample, we use the `nmr_align` function, which in turn uses a hierarchical
clustering method (see `?nmr_align` for further details).

The `maxShift_ppm` limits the maximum shift allowed for the spectra.


```{r}
NMRExp_ref <- nmr_align_find_ref(dataset, peak_list)
message("Your reference is NMRExperiment ", NMRExp_ref)
```


```{r}
dataset_align <- nmr_align(dataset, peak_list, NMRExp_ref, maxShift_ppm = 0.0015, acceptLostPeak = FALSE)
```

```{r}
plot(dataset, chemshift_range = c(3.025, 3.063))
plot(dataset_align, chemshift_range = c(3.025, 3.063))
```

```{r}
cowplot::plot_grid(
    plot(dataset, chemshift_range = c(3.22, 3.25)) + theme(legend.position = "none"),
    plot(dataset_align, chemshift_range = c(3.22, 3.25)) + theme(legend.position = "none")
)
```


# Normalization

There are multiple normalization techniques available. The most strongly 
recommended is the `pqn` normalization, but it may not be fully reliable
when the number of samples is small, as it needs a  computation of the 
median spectra. Nevertheless, it is possible to compute it:

```{r}
dataset_norm <- nmr_normalize(dataset_align, method = "pqn")
```

The `AlpsNMR` package offers the possibility to extract additional 
normalization information with `nmr_normalize_extra_info(dataset)`, to explore
the normalization factors applied to each sample:

The plot shows the dispersion with respect to the median of the normalization
factors, and can highlight samples with abnormally large or small normalization
factors.

```{r}
diagnostic <- nmr_normalize_extra_info(dataset_norm)
diagnostic$norm_factor
diagnostic$plot
```


# Peak integration

## 1. Integration based on peak center and width

If we want to integrate the whole spectra, we need ppm from the `peak_table`. 
See `Peak detection` section. The function `nmr_integrate_peak_positions`
generates a new `nmr_dataset_1D` object containing the integrals from 
the `peak_table` (ppm values corresponding to detected peaks).

```{r}
peak_table_integration = nmr_integrate_peak_positions(
  samples = dataset_norm,
  peak_pos_ppm = peak_list$ppm[peak_list$NMRExperiment == NMRExp_ref],
  peak_width_ppm = 0.006)

peak_table_integration = get_integration_with_metadata(peak_table_integration)
```


We can also integrate with a specific peak position and some arbitrary width:

```{r}
nmr_data(
  nmr_integrate_peak_positions(samples = dataset_norm,
                            peak_pos_ppm = c(4.1925, 4.183, 4.1775, 4.17),
                            peak_width_ppm = 0.006)
)
```


## 2. Integration based on peak boundaries

Imagine we only want to integrate the four peaks corresponding to the pyroglutamic
acid:

```{r}
pyroglutamic_acid_region <- c(4.15, 4.20)
plot(dataset_norm, chemshift_range = pyroglutamic_acid_region) +
  ggplot2::ggtitle("Pyroglutamic acid region")
```

We define the peak regions and integrate them. Note how we can correct the
baseline or not. If we correct the baseline, the limits of the integration will
be connected with a straight line and that line will be used as the baseline,
that will be subtracted.

```{r}
pyroglutamic_acid <- list(pyroglutamic_acid1 = c(4.19, 4.195),
                          pyroglutamic_acid2 = c(4.18, 4.186),
                          pyroglutamic_acid3 = c(4.175, 4.18),
                          pyroglutamic_acid4 = c(4.165, 4.172))
regions_basel_corr_ds <- nmr_integrate_regions(dataset_norm, pyroglutamic_acid, fix_baseline = TRUE)
regions_basel_corr_matrix <- nmr_data(regions_basel_corr_ds)
regions_basel_corr_matrix


regions_basel_not_corr_ds <- nmr_integrate_regions(dataset_norm, pyroglutamic_acid, fix_baseline = FALSE)
regions_basel_not_corr_matrix <- nmr_data(regions_basel_not_corr_ds)
regions_basel_not_corr_matrix
```

We may plot the integral values to explore variation based on the baseline
subtraction.

```{r}
dplyr::bind_rows(
  regions_basel_corr_matrix %>%
    as.data.frame() %>%
    tibble::rownames_to_column("NMRExperiment") %>%
    tidyr::gather("metabolite_peak", "area", -NMRExperiment) %>%
    dplyr::mutate(BaselineCorrected = TRUE),
  regions_basel_not_corr_matrix %>%
    as.data.frame() %>%
    tibble::rownames_to_column("NMRExperiment") %>%
    tidyr::gather("metabolite_peak", "area", -NMRExperiment) %>%
    dplyr::mutate(BaselineCorrected = FALSE)
) %>% ggplot() + geom_point(aes(x = NMRExperiment, y = area, color = metabolite_peak)) +
  facet_wrap(~BaselineCorrected)

```


# Identification

After applying any feature selection or machine learning, Alps allows the 
identification of features of interest through `nmr_identify_regions_blood`. 
The function gives 3 posibilities sorted by the most probable metabolite 
(see `nmr_identify_regions_blood` for details).

```{r}
ppm_to_assign <- c(4.060960203, 3.048970634,2.405935596,0.990616851,0.986520147, 1.044258467)
identification <- nmr_identify_regions_blood (ppm_to_assign)
```


# Free experimentation

## Getting the spectra and manipulating it manually

Besides all those techniques, you can easily implement your own. You can extract
the raw matrix and manipulate it at will. As long as you don't
permute the rows, you can always replace the raw matrix of the `nmr_dataset_1D`
object through the `nmr_data` function:

```{r}
full_spectra_matrix <- nmr_data(dataset)
full_spectra_matrix[1:3, 1:6] # change it as you wish
nmr_data(dataset) <- full_spectra_matrix # Rewrite the matrix
```


## Creating an `nmr_dataset_1D` object from a matrix

You can also create an `nmr_dataset_1D` object from scratch with the
`new_nmr_dataset_1D` function:

```{r}
nsamp <- 12
npoints <- 20
# Create a random spectra matrix
dummy_ppm_axis <- seq(from = 0.2, to = 10, length.out = npoints)
dummy_spectra_matrix <- matrix(runif(nsamp*npoints), nrow = nsamp, ncol = npoints)
metadata <- list(external = data.frame(NMRExperiment = paste0("Sample", 1:12),
                                       DummyClass = c("a", "b"),
                                       stringsAsFactors = FALSE))
your_custom_nmr_dataset_1D <- new_nmr_dataset_1D(ppm_axis = dummy_ppm_axis,
                                                 data_1r = dummy_spectra_matrix,
                                                 metadata = metadata)
your_custom_nmr_dataset_1D
plot(your_custom_nmr_dataset_1D) +
  ggtitle("Of course those random values don't make much sense...")
```


# Final thoughts

This vignette shows many of the features of the package, some features have
room for improvement, others are not fully described, and the reader will need
to browse the documentation. Hopefully it is a good starting point for using the
package.

```{r}
sessionInfo()
```
