---
title: "From the peak table to the differential analysis"
date: "`r format(Sys.Date(), '%F')`"
output:
  BiocStyle::pdf_document
vignette: >
  %\VignetteIndexEntry{Vignette 04: From peak table to differential analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(AlpsNMR)
```



To be written.

# QC on the peak table

- Plots
- PCA
- ...

# Univariate analysis

- hypothesis testing

# Multivariate analysis


```
nmr_data_analysis
```

# Identification of relevant features

After applying any feature selection or machine learning, Alps allows the 
identification of features of interest through `nmr_identify_regions_blood`. 
The function gives 3 posibilities sorted by the most probable metabolite 
(see `nmr_identify_regions_blood` for details).

```{r}
ppm_to_assign <- c(4.060960203, 3.048970634,2.405935596,0.990616851,0.986520147, 1.044258467)
identification <- nmr_identify_regions_blood(ppm_to_assign)
identification
```


# Session Info:

```{r}
sessionInfo()
```
